\textbf{Generalized additive models to analyze biomedical non-linear longitudinal data in R: Beyond repeated measures ANOVA and Linear Mixed Models}

### 2021-11-17 Manuscripts/Manuscript_by_chapters-SIM_Revisions/Full_document_SIM_No_Appendix.Rmd
Updated figures and theme using scico

#global themes for plotting
#thm<-scale_color_viridis_d(option="turbo",begin=0,end=1)
#thm1<-scale_fill_viridis_d(option="turbo",begin=0,end=1)#global theme for plotting
thm1<-scale_fill_scico_d(palette="oslo")

>>>>

#global themes for plotting
#thm<-scale_color_viridis_d(option="turbo",begin=0,end=1)
#thm1<-scale_fill_viridis_d(option="turbo",begin=0,end=1)#global theme for plotting
thm1<-scale_fill_scico_d(palette="tokyo",begin=0.2, end=0.8, direction = -1, aesthetics = c("colour","fill"))

### 2021-11-17 Manuscripts/Manuscript_by_chapters-SIM_Revisions/Full_document_SIM_No_Appendix.tex
Updated figures and theme using scico

The simulated data is based on the reported longitudinal changes in oxygen saturation (\(\mbox{StO}_2\)) in subcutaneous tumors that appear in Figure 3C in.\textsuperscript{\protect\hyperlink{ref-vishwanath2009}{16}} In the paper, diffuse reflectance spectroscopy was used to quantify \(\mbox{StO}_2\) changes in both groups at the same time points (days 0, 2, 5, 7 and 10). In the ``Treatment'' group (chemotherapy) an increase in \(\mbox{StO}_2\) is observed through time, while a decrease is seen in the ``Control'' (saline) group. Following the reported trend, we simulated 10 normally distributed observations at each time point with a standard deviation (SD) of 10\% (matching the SD in the original paper).
The simulated and real data appear in Figure \ref{fig:sim-smooth-plot},A.

>>>>

The simulated data is based on the reported longitudinal changes in oxygen saturation (\(\mbox{StO}_2\)) in subcutaneous tumors (Figure 3C in Vishwanath et. al.\textsuperscript{\protect\hyperlink{ref-vishwanath2009}{16}}). In the paper, diffuse reflectance spectroscopy was used to quantify \(\mbox{StO}_2\) changes in both groups at the same time points (days 0, 2, 5, 7 and 10). In the ``Treatment'' group (chemotherapy) an increase in \(\mbox{StO}_2\) is observed through time, while a decrease is seen in the ``Control'' (saline) group. Following the reported trend, we simulated 10 normally distributed observations at each time point with a standard deviation (SD) of 10\% (matching the SD in the original paper).
The simulated and real data appear in Figure \ref{fig:sim-smooth-plot},A.

### 2021-11-17 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/03-GAM Theory.Rmd
Updated figures and theme using scico

#plotting simulated datapoints and smooth term
p14<-ggplot(data=dat,
            aes(x=time,y=y))+
  geom_point(size=sz+1)+
  thm1+
  labs(y='Simulated \n response')+
  geom_line(data=smooth,
            aes(x=time,
                y=smooth),
            color="#6C581DFF",
            size=sz+1)+
  theme_classic()

>>>>

#plotting simulated datapoints and smooth term
p14<-ggplot(data=dat,
            aes(x=time,y=y))+
  geom_point(size=sz+1,alpha=0.5)+
  thm1+
  labs(y='Simulated \n response')+
  geom_line(data=smooth,
            aes(x=time,
                y=smooth),
            color="#6C581DFF",
            size=sz+1)+
  theme_classic()

### 2021-11-17 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/04-Longitudinal analysis with GAMs.Rmd
Updated figures and theme using scico

```{r,sim-tumor-response,echo=FALSE,include=FALSE,message=FALSE,warning=FALSE}
dat<-tibble(StO2=c(4,27,3,2,0.5,7,4,50,45,56),
            Day=rep(c(0,2,5,7,10),times=2),
            Group=as.factor(rep(c("Control","Treatment"),each=5))
            )

## plot the mean response
f1<-ggplot(dat, 
           aes(x = Day, 
               y = StO2, 
               color = Group)) +
    geom_line(size=1,
              show.legend = FALSE)+
    geom_point(show.legend = FALSE,
               size=1.5,
               alpha=0.5)+
  labs(y=expression(paste(StO[2],
                          ' (real)')))+
  theme_classic()+
  thm1+
    scale_x_continuous(breaks=c(0,5,10))+
    scale_y_continuous(breaks=c(0,40))+
  plot_layout(tag_level = 'new')+
  theme(
    plot.background = element_rect(fill = "transparent", 
                                   color = NA),
    axis.text=element_text(size=14)
  )

@@

#plotting simulated data
f2<-ggplot(dat_sim, aes(x = Day, 
                        y = StO2_sim, 
                        color = Group, 
                        group=subject)) +
    geom_point(show.legend=FALSE,
               size=1.5,
               alpha=0.6)+
    geom_line(size=0.6, alpha=0.4)+
    geom_line(aes(x = Day, 
               y = StO2,
               color=Group),
           size=1.5,
           data=dat,
           inherit.aes=FALSE,
              show.legend = FALSE)+
  labs(y=expression(atop(StO[2],'(simulated)')))+
  theme_classic()+
  theme(
    axis.text=element_text(size=22)
  )+
  thm1+
    scale_x_continuous(breaks=c(0,2,5,7,10))

@@

#plot smooths and confidence interval for GAM
f3<-ggplot(data=dat_sim, aes(x=Day, y=StO2_sim, group=Group)) +
    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
                   ymax=(fit + 2*se.fit),
                   fill=Group
                   ),
              alpha=0.5,
              data=gam_predict,
            show.legend=FALSE,
            inherit.aes=FALSE) +
  geom_line(aes(y=fit,
                color=Group),
              size=1,data=gam_predict,
              show.legend = FALSE)+
  #facet_wrap(~Group)+
  labs(y=expression(atop(StO[2],'complete')))+
    scale_x_continuous(breaks=c(0,2,5,7,10))+
      theme_classic()+
  theme(
    axis.text=element_text(size=22)
  )

#plot linear fit for rm-ANOVA
f4<-ggplot(data=dat_sim, aes(x=Day, y=StO2_sim, group=Group)) +
    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
                   ymax=(fit + 2*se.fit),fill=Group),
              alpha=0.5,
              data=lm_predict,
              show.legend = FALSE,
                inherit.aes=FALSE) +
  geom_line(aes(y=fit,
                color=Group),
              size=1,data=lm_predict,
              show.legend = FALSE)+
  #facet_wrap(~Group)+
  labs(y=expression(paste('StO'[2],' (simulated)')))+
    scale_x_continuous(breaks=c(0,2,5,7,10))+
      theme_classic()+
  theme(
    axis.text=element_text(size=22)
  )

@@

f6<-ggplot(data=dat_missing, aes(x=Day, y=StO2_sim, group=Group)) +
    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
                   ymax=(fit + 2*se.fit),
                   fill=Group
                   ),
              alpha=0.5,
              data=m_predict,
            show.legend=FALSE,
                inherit.aes=FALSE) +
  geom_line(aes(y=fit,
                color=Group),
              size=1,data=m_predict,
              show.legend = TRUE)+
  #facet_wrap(~Group)+
  labs(y=expression(atop(StO[2],'missing')))+
    scale_x_continuous(breaks=c(0,2,5,7,10))+
      theme_classic()+
  theme(
    axis.text=element_text(size=22)
  )

>>>>

```{r,sim-tumor-response,echo=FALSE,include=FALSE,message=FALSE,warning=FALSE}
dat<-tibble(StO2=c(4,27,3,2,0.5,7,4,50,45,56),
            Day=rep(c(0,2,5,7,10),times=2),
            Group=as.factor(rep(c("Control","Treatment"),each=5))
            )
#alpha for ribbon
al<-0.8

## plot the mean response
f1<-ggplot(dat, 
           aes(x = Day, 
               y = StO2, 
               color = Group)) +
    geom_line(size=1,
              show.legend = FALSE)+
    geom_point(show.legend = FALSE,
               size=1.5,
               alpha=0.5)+
  labs(y=expression(paste(StO[2],
                          ' (real)')))+
  theme_classic()+
    scale_x_continuous(breaks=c(0,5,10))+
    scale_y_continuous(breaks=c(0,40))+
  plot_layout(tag_level = 'new')+
  theme(
    plot.background = element_rect(fill = "transparent", 
                                   color = NA),
    axis.text=element_text(size=14)
  )+
    thm1

@@

#plotting simulated data
f2<-ggplot(dat_sim, aes(x = Day, 
                        y = StO2_sim, 
                        color = Group, 
                        group=subject)) +
    geom_point(show.legend=FALSE,
               size=1.5,
               alpha=0.6)+
    geom_line(size=0.6, alpha=0.4)+
    geom_line(aes(x = Day, 
               y = StO2,
               color=Group),
           size=1.5,
           data=dat,
           inherit.aes=FALSE,
              show.legend = FALSE)+
  labs(y=expression(atop(StO[2],'(simulated)')))+
  theme_classic()+
  theme(
    axis.text=element_text(size=22)
  )+
    scale_x_continuous(breaks=c(0,2,5,7,10))+
    thm1

@@

#plot smooths and confidence interval for GAM
f3<-ggplot(data=dat_sim, aes(x=Day, y=StO2_sim, group=Group)) +
    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
                   ymax=(fit + 2*se.fit),
                   fill=Group
                   ),
              alpha=al,
              data=gam_predict,
            show.legend=FALSE,
            inherit.aes=FALSE) +
  geom_line(aes(y=fit,
                color=Group),
              size=1,data=gam_predict,
              show.legend = FALSE)+
  #facet_wrap(~Group)+
  labs(y=expression(atop(StO[2],'complete')))+
    scale_x_continuous(breaks=c(0,2,5,7,10))+
      theme_classic()+
  theme(
    axis.text=element_text(size=22)
  )+thm1

#plot linear fit for rm-ANOVA
f4<-ggplot(data=dat_sim, aes(x=Day, y=StO2_sim, group=Group)) +
    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
                   ymax=(fit + 2*se.fit),fill=Group),
              alpha=al,
              data=lm_predict,
              show.legend = FALSE,
                inherit.aes=FALSE) +
  geom_line(aes(y=fit,
                color=Group),
              size=1,data=lm_predict,
              show.legend = FALSE)+
  #facet_wrap(~Group)+
  labs(y=expression(paste('StO'[2],' (simulated)')))+
    scale_x_continuous(breaks=c(0,2,5,7,10))+
      theme_classic()+
  theme(
    axis.text=element_text(size=22)
  )+thm1

@@

f6<-ggplot(data=dat_missing, aes(x=Day, y=StO2_sim, group=Group)) +
    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
                   ymax=(fit + 2*se.fit),
                   fill=Group
                   ),
              alpha=al,
              data=m_predict,
            show.legend=FALSE,
                inherit.aes=FALSE) +
  geom_line(aes(y=fit,
                color=Group),
              size=1,data=m_predict,
              show.legend = TRUE)+
  #facet_wrap(~Group)+
  labs(y=expression(atop(StO[2],'missing')))+
    scale_x_continuous(breaks=c(0,2,5,7,10))+
      theme_classic()+
  theme(
    axis.text=element_text(size=22)
  )+thm1
